---
title: "Unit 3 | Advanced Correlation Analysis"
author: "FIRSTNAME LASTNAME"
date: "`r format(Sys.time(), 'Compiled %X on %A, %B %d, %Y')`"
output: pdf_document
---

# Load packages

```{r load packages, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Load packages
library(psych); library(polycor); library(GGally); library(ggplot2); library(ppcor)
```

# Motivating Dataset 1: CERN

Research question: For particles with very little mass, can we predict their energy given momentum alone?

$$
H_0: \rho_{e, m} = 0
$$

$$
H_A: \rho_{e, m} \neq 0
$$

## Load in the Data

```{r load cern data, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Load CERN data
cern <- read.csv("data/cern.csv")
```

## Your Turn: Wrangle the Data

1.    Subset to only include `Run` 165617

2.    Select variables of interest: `E1`, `px1`, `py1`, `pz1`

3.    Name final object as `cern_voi` (variables of interest)

```{r wrangle cern data, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}

```

## Visualize Dataset using `ggpairs`

```{r ggpairs, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Visualize
ggpairs(cern_voi)
```

## Non-graphical Correlations

```{r correlation, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Correlation between two variables
cor(cern_voi$E1, cern_voi$px1)

# Correlation test between two variables
cor.test(cern_voi$E1, cern_voi$px1)
```

> Interpret this test: Is it significant? How many cases are in the sample (hint: look at `df`)

```{r correlation matrix, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Correlation matrix
round(cor(cern_voi), 3)
```

## Your Turn: Create 3D Momentum

Compute 3D momentum by taking the square root of the sum of the squares of these three variables: `px1`, `py1`, `pz1`

1.    Square the variables

2.    Sum the squared values

3.    Take the square root of the sum

This computation *can* be done in a single line of code! (*Hint*: you'll need to sum the rows!)

Store the values as `p1` in your `cern_voi` data frame

```{r 3D momentum, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}

```

Now, plot `E1` and `p1` as a scatterplot using \{ggplot2\}

```{r 3D momentum scatter, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}

```

> What do you think the correlation value is (without performing `cor`)?

# Motivating Dataset 2: Rate My Professor

## Load in the Data

```{r load rate my professor data, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Load rate my professor data
rmp <- read.csv("./data/ratemyprofessor.csv")

colnames(rmp)[31:50]
```

> Come up with a research question that looks at the association between two variables (write it down below)

> Write down the formal hypothesis test

$$

$$

$$

$$

> Predict what you think the relationship will be (small, moderate, large; positive/negative, significant/non-significant)

+   small $r$ = 0.10

+   moderate $r$ = 0.30

+   large $r$ = 0.50

Prediction:

<!---

Do not advance to next step until you have a research question and formal hypothesis about the strength and direction of the effect!

If you create a research question/hypothesis after knowing the results, that's extremely frowned upon!

HARKing -- *H*ypothesizing *A*fter *R*esults are *K*nown

--->

## Perform the tetrachoric correlation

```{r compute tetrachoric correlation, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Tetrachoric correlation
output <- polycor::polychor(
  x = , # one variable here
  y = , # another variable here
  std.err = TRUE # gets p-value (in parenthesis after coefficient)
)
```

The *p*-value is not given to us automatically, so we need to compute it using the standard error.

```{r compute p value, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Compute z-statistic: correlation divided by standard error
z_stat <- output$rho / sqrt(output$var)

# Compute p-value
2 * pnorm(abs(z_stat), lower.tail = FALSE)
```

> Interpret the result (what is the correlation and *p*-value?)

> Was your hypothesis correct?

# Motivating Dataset 3: Life Expectancy

Research question: What is the predictive impact of population levels of substance use on life expectancy *given* GDP?

*"substance use" will be defined as: per-capita alcohol consumption*

> Reflection: Is per-capita alcohol consumption an adequate indicator of "substance use"? Is that how *you* would define substance use?

## Load in the Data

```{r load life expectancy data, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Load life expectancy data
life_expectancy <- read.csv("./data/life_expectancy.csv")
```

## Your Turn: Visualize Columns *4-13* using `ggpairs`

```{r ggpairs life, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}

```

## Your Turn: Perform `pcor.test` on the necessary variables to answer the research question
*Hint*: you'll need to remove a missing data point

```{r , echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Partial correlation test

```

> Interpret the output

> What is the correlation? Is it small, moderate, or large? Positive or negative?

> What is the *p*-value? Is it significant?

> If you were to tell someone about this relationship who doesn't know anything about statistics, what would you say?
