---
title: "Statistics Refresher, Part 1: The p-Value and Power"
author: "YOUR NAME HERE"
date: "DS 3100 --- Thursday 24 September 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## What you are doing

On Thursday we ended the work on figures with a question. The women in
the smartphone survey scored about 12 points higher than the men on
problematic phone use, and the figure's title says so. That is true of
these 121 students. Would a difference of 12 points appear in a sample
even if there were no difference in the population? This sheet builds
the answer (a p-value) by simulation before `t.test()` computes it from
a formula, then asks how often a test finds a difference that is real,
which is the test's power. **Part 1** is four sections we run in class
with the slides. **Part 2** is three investigations to run on your own:
each is a machine with dials, and there is no answer on the sheet.

Every chunk of code is switched off with a hash character at the start
of each line. Please remove the hashes when you reach a chunk (select
the lines and press Ctrl+Shift+C, or Cmd+Shift+C on a Mac), run it, and
look at the result before reading on. After each section, there is a
question I want you to answer in your own words.

The data are the smartphone survey we have used since 10 September
(`PluckData.rds`, attached again on Brightspace and linked from the
course website). Create a folder called `data/` beside this file and
put the file in it.

```{r load-packages}
library(tidyverse)
```

The first chunk reads the file and makes the fix from 15 September:
three students have a GPA of 999, a code for a missing value, which
`na_if()` turns into `NA`. `by_sex` keeps the 121 students whose sex
and phone-use score (`MPPUS__Total`, the Mobile Phone Problem Use
Scale) are both recorded.

```{r read-data}
#pluck <- readRDS("data/PluckData.rds") |>
#  mutate(GPA = na_if(GPA, 999))
#by_sex <- pluck |> filter(!is.na(Sex), !is.na(MPPUS__Total))
#nrow(by_sex)
```

`nrow()` should return 121.


# Part 1: In class

`set.seed()` fixes the starting point of R's random numbers, so that
if you run the chunks of Part 1 in order, once each, your numbers match
the ones on the slides. Every run after the first gives different
numbers, which is the variation from sample to sample that this part
is about.

## 1. A real difference, or chance?

The women's mean is about 12 points higher than the men's.

```{r observed}
#set.seed(3100)
#observed <- mean(by_sex$MPPUS__Total[by_sex$Sex == "Female"]) -
#            mean(by_sex$MPPUS__Total[by_sex$Sex == "Male"])
#observed
```

Suppose sex had nothing to do with the score. Then the labels "Male"
and "Female" might as well have been dealt out at random, and any
dealing is as likely as the real one. `sample(by_sex$Sex)` shuffles the
labels. `replicate()` runs the code in its braces ten thousand times
and collects the ten thousand results: here, the difference between
the two means after each shuffle. That is what a difference of two
means looks like when the truth is "no difference".

Before you run the next chunk, please write down how often you expect
a shuffle to give a difference of 12 points or more.

> My prediction:

```{r shuffle}
#shuffled <- replicate(10000, {
#  s <- sample(by_sex$Sex)
#  mean(by_sex$MPPUS__Total[s == "Female"]) - mean(by_sex$MPPUS__Total[s == "Male"])
#})
#hist(shuffled, breaks = 40)
#abline(v = c(-observed, observed), col = "red", lwd = 2)
#mean(abs(shuffled) >= abs(observed))
```

The last line is a p-value: the proportion of the shuffles in which
chance alone produced a difference at least as large as the real one,
in either direction. `t.test()` computes the same thing from a formula
instead of from shuffling. Please compare its `p-value` with yours.

```{r t-test}
#t.test(MPPUS__Total ~ Sex, data = by_sex)
```

Please write, in one sentence, what your p-value means. Then write one
thing it does *not* mean.

> Answer:


## 2. The false-positive machine

`rnorm(30)` draws thirty values from a normal population with mean 0
and standard deviation 1. It is a machine that makes data, and we know
exactly what it is doing, which is never true of real data. Here two
samples of thirty are drawn from the *same* population, so there is no
difference to find, and a t-test is run on them. `$p.value` takes the
p-value out of the test's output. The chunk does this a thousand times
and keeps the thousand p-values.

Before you run it, please write down how many of the thousand tests
you expect to come out below 0.05.

> My prediction:

```{r false-positives}
#p_null <- replicate(1000, t.test(rnorm(30), rnorm(30))$p.value)
#hist(p_null, breaks = 20)
#sum(p_null < 0.05)
#mean(p_null < 0.05)
```

What shape does the histogram of p-values have when there is nothing
to find? What does a single result of p = 0.03 tell you on its own?

> Answer:


## 3. Power: how often a real difference is found

Now the machine is given a real difference. `rnorm(n, 0.5)` has mean
0.5 and `rnorm(n)` has mean 0, so the two groups differ by half a
standard deviation. `power_sim()` runs the test two thousand times and
returns the proportion of tests that found the difference (p below
0.05). That proportion is the test's **power** at that sample size.

```{r power}
#power_sim <- function(n, d, reps = 2000) {
#  mean(replicate(reps, t.test(rnorm(n, d), rnorm(n))$p.value < 0.05))
#}
#power_sim(10, 0.5)
#power_sim(30, 0.5)
#power_sim(64, 0.5)
#power_sim(100, 0.5)
```

`power.t.test()` does the same calculation from a formula, and it can
answer the question the other way round: how many people per group
for a power of 80 percent?

```{r power-formula}
#power.t.test(n = 10, delta = 0.5, sd = 1)
#power.t.test(delta = 0.5, sd = 1, power = 0.8)
```

A study with ten people per group reports "no significant difference".
Is that evidence that there is no difference? Please explain using the
numbers you got.

> Answer:


## 4. Power for our own question

Suppose the true difference in the population were exactly the one in
the survey, about 12 points, with the spread the survey shows. What
power did the survey have to find it? `power.t.test()` assumes two
groups of the same size, and the survey has 51 men and 70 women; two
groups of 59 give about the same precision (59 is the harmonic mean,
`2 / (1/51 + 1/70)`). It uses `by_sex` and `observed` from the start of
section 1, so in a new session please run the chunks from the top first.

```{r room-power}
#spread <- sd(by_sex$MPPUS__Total)
#spread
#power.t.test(n = 59, delta = observed, sd = spread)
#power.t.test(delta = observed, sd = spread, power = 0.8)
```

What power did the survey have, and how many people per group would
it take to reach 80 percent? What does that tell you about the p-value
from section 1?

> Answer:

# Part 2: On your own

Part 2 is for after class. Each investigation is a machine with dials,
and there is no answer on the sheet. Please turn the dials, watch what
happens, and write in two or three sentences what you concluded and
what you saw that made you conclude it. Please bring your conclusions
to Tuesday.


## Investigation 1: the shape of a thousand p-values

Section 2 drew the histogram of p-values when there was nothing to
find. This machine adds a real difference of size `d` (in standard
deviations) and draws the histogram again. Please try `d` at 0, 0.2,
0.5, and 1, and `n` at 20 and 100.

```{r investigation-1}
#p_hist <- function(n, d, reps = 1000) {
#  p <- replicate(reps, t.test(rnorm(n, d), rnorm(n))$p.value)
#  hist(p, breaks = 20, main = paste("n =", n, "  d =", d))
#  mean(p < 0.05)
#}
#p_hist(20, 0)
#p_hist(20, 0.5)
```

If someone handed you a histogram of the p-values from a thousand
studies of one question, what could you tell about whether the effect
is real, and about how big it is?

> Conclusion:


## Investigation 2: how many discoveries are false?

Statement 5 from class said that after rejecting the null hypothesis
we know the probability that we are wrong. This machine shows what
that probability depends on. Imagine ten thousand hypotheses being
tested across a field. A fraction `prior` of them are actually true. A
test of a true hypothesis comes out significant with probability
`power`, and a test of a false one comes out significant with
probability 0.05 (the false-positive rate from section 2). The machine
counts, among the significant results, the proportion that were true.

```{r investigation-2}
#discoveries <- function(prior, power, n_tests = 10000, alpha = 0.05) {
#  is_true <- rbinom(n_tests, 1, prior) == 1
#  significant <- ifelse(is_true, rbinom(n_tests, 1, power), rbinom(n_tests, 1, alpha)) == 1
#  sum(is_true & significant) / sum(significant)
#}
#discoveries(prior = 0.1, power = 0.5)
```

Please try `prior` at 0.5, 0.1, and 0.01, and `power` at 0.8, 0.5, and
0.2, in every combination that interests you.

Under what conditions is a significant result more likely to be wrong
than right, and what does that say about fields where most of the
hypotheses tested are long shots?

> Conclusion:


## Investigation 3: significant, and worth nothing

A real difference of 0.05 standard deviations (a twentieth of one) is
about as small as an effect can be and still exist. The machine tests
it with ten thousand people per group. It uses `power_sim()` from
section 3, so please run that chunk first.

```{r investigation-3}
#power_sim(10000, 0.05, reps = 200)
#t.test(rnorm(10000, 0.05), rnorm(10000))
```

Please compare with `power_sim(30, 0.05)`, and look at the `sample
estimates` and the `95 percent confidence interval` lines of the
`t.test()` output, not only the p-value.

What does a p-value tell you about the *size* of an effect, and what
part of the output would you report instead?

> Conclusion:

## When you have finished

Knit this document (HTML is fine). If any chunk printed something you
did not expect, please write one sentence under it saying what you
expected instead. Please bring your conclusions from Part 2 to
Tuesday's meeting (in Stevenson 5, Room 306), where the refresher
continues.
