---
title: "Stats Intermission"
author: "FIRSTNAME LASTNAME"
date: "`r format(Sys.time(), 'Compiled %X on %A, %B %d, %Y')`"
output: pdf_document
---

# Load packages

```{r, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Load packages
library(pwr); library(MASS); library(psych)
```

Effect size

+   small: $r = 0.10$

+   moderate: $r = 0.30$

+   large: $r = 0.50$

+   very large: $r = 0.70$


# Power, Effect Size, and Significance

```{r, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Test up to one hundred thousand
sample_sizes <- c(100, 1000, 10000, 100000)

# Initialize data frame
result <- data.frame(
  N = sample_sizes,
  effect = numeric(length(sample_sizes)),
  p_value = numeric(length(sample_sizes))
)

# Populate data frame
for(i in seq_along(sample_sizes)){
  
  # Re-set seed (for reproducibility)
  set.seed(43)
  
  # Generate data
  data <- mvrnorm(
    n = sample_sizes[i], mu = c(0, 0), 
    Sigma = matrix(c(1, 0.01, 0.01, 1), nrow = 2)
  ) 

  
  # Get correlation
  r <- cor.test(data[,1], data[,2])
  
  # Populate data frame
  result$effect[i] <- r$estimate
  result$p_value[i] <- r$p.value
  
}

# Print
print(result, digits = 3)
```

# Starting with Hypotheses

## Formal Notation

$$
H_0: \rho_{x, y} = 0
$$

<!--

LaTeX math notation

single dollar signs "$ formula $" = in line equation

double dollar signs "$$ formula $$" = new paragraph equation

"_" = creates a subscript

"{}" = include "everything" in whatever you're doing, so "_{x, y}" creates a subscript with x comma y

Resource: https://en.wikibooks.org/wiki/LaTeX/Mathematics

-->

$$
H_A: \rho_{x, y} \neq 0
$$

<!--

LaTeX math symbols

"\neq" = does not equal symbol

"\alpha" = lowercase greek letter "a"

The "\" backslash denotes a special behavior for the text -- rather than writing the text it tells LaTeX to do something different

Resource: https://www.cmor-faculty.rice.edu/~heinken/latex/symbols.pdf

-->

## Interpretation of formal notation

$H_0: \rho_{x,y} = 0$: The null hypothesis is that the correlation between variable $x$ and $y$ equals zero

$H_A: \rho_{x,y} \neq 0$: The alternative hypothesis is that the correlation between variable $x$ and $y$ does *not* equal zero

## Two mean comparison

What about for the difference between two means?

Null hypothesis: The difference of the mean of variables $x$ and $y$ equals zero

$$
H_0: \mu_x - \mu_y = 0
$$

Alternative hypothesis: The difference of the mean of variables $x$ and $y$ does *not* equal zero

$$
H_A: \mu_x - \mu_y \neq 0
$$
## Statistical Significance

$$
\alpha = 0.05
$$

Statistical significance threshold

$$
p < 0.05
$$

This value means we win at science, right?

<!--

Definition of p-value: The probability of obtaining a test statistic as extreme or more extreme given the null hypothesis is true

-->

## Your Turn: run this code

```{r, comment = NA, warning = FALSE, message = FALSE, echo = TRUE, eval = FALSE}
# Perform a correlation test
cor.test(rnorm(100), rnorm(100))
```

> Write down your p-value: 

> Send your p-value into Top Hat so we can compare as a class

## Demonstrating false positive rates

```{r, comment = NA, warning = FALSE, message = FALSE, echo = TRUE, eval = TRUE}
set.seed(986098)
# Perform 1000 correlation tests and save out p-values
pvals <- data.frame() #create an empty dataframe to store p-values

for(i in 1:100){
  pvals[i,1] <- cor.test(rnorm(100), rnorm(100))$p.value #save out p-values
  pvals[i, 2] <- ifelse(pvals[i,1] < .05, 1, 0) #keep track of significant results
}

describe(pvals)
```

# The Interaction between Power

## Finding Power

Using the \{pwr\} package, compute power for the correlation of effect size of 0.01 at sample sizes of 10, 100, 1000, and 100000

```{r correlation power, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# N = 10
pwr.r.test(n = 10, r = 0.50)

# N = 100
pwr.r.test(n = 100, r = 0.50)

# N = 1000
pwr.r.test(n = 1000, r = 0.50)

# N = 100000
pwr.r.test(n = 100000, r = 0.50)

```

> What happens to power as sample size increases?

## Finding Sample Size

Now, rather than finding the power, find the sample size needed for a power of 0.80 and correlation effect size of 0.01

```{r correlation sample size, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Power of 0.80 
pwr.r.test(
  n = NULL, # sample size
  r = 0.10, # effect size
  power = 0.80, # statistical power
  sig.level = 0.05 # statistical significance
)
# You should *always* leave the metric you want missing (NULL)!!
```

## Finding Effect Size

Finally, what effect size do you have the power for when power is 0.80 and sample size is 10, 100, 1000, and 100000?

```{r correlation effect size, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# N = 10

# N = 100

# N = 1000

# N = 100000

```

> Reflection: Based on our correlation example from before, how does power relate to our findings?