---
title: "Lab 4: Interpretations and Discussion"
author: "FIRSTNAME LASTNAME"
date: "`r format(Sys.time(), 'Compiled %X on %A, %B %d, %Y')`"
output: pdf_document
---

<!--

These comments contain instructions and do not appear in the knitted
document. You may delete them after completing the steps or leave
them in place.

This lab runs in section. The completed sheet, its PDF, and the data
files it reads go into the Lab 4 folder on Brightspace before you
leave.

Before you start: rename this file Lab-4_FIRSTNAME_LASTNAME.Rmd, put
your name in the YAML header above (keep the quotation marks), and put
nashville_daily.rds and your own data file in a data/ folder beside
this file. The R code is commented out; remove the hash characters at the
start of code lines once the files are in place.

The three parts follow the handout. Part 1 computes the numbers
behind two of the Nashville figures and writes their interpretations,
Part 2 writes a Discussion for the Nashville question, and Part 3 does
both for your own report.

-->

```{r setup, include = FALSE}
library(tidyverse)
```

# Part 1: The numbers behind a figure

<!-- The handout's research question: Are Nashville's summers getting
     hotter, and is it the whole summer or only its hottest days? The
     three figures are in the handout. This part redraws Figure 2 and
     computes the numbers behind Figures 2 and 3. -->

## The data and the two periods

The file has one row per day at Nashville International Airport from
1991 to 2026. `tmax_f` is the day's high in degrees Fahrenheit. The
chunk keeps the summer days (June to August) of the two periods the
figures compare, and splits them into two tibbles, one per period.

```{r read-nashville}
#nashville <- readRDS("data/nashville_daily.rds")
#summers <- nashville |>
#  filter(month %in% 6:8) |>
#  mutate(period = case_when(year <= 2005 ~ "1991-2005",
#                            year >= 2012 ~ "2012-2026")) |>
#  filter(!is.na(period))
#early <- filter(summers, period == "1991-2005")
#late  <- filter(summers, period == "2012-2026")
```

What happens to the summers of 2006 to 2011 in this chunk, and why
might the figures leave them out?

*Your answer here.*

## Figure 2: the shape of the daily highs

A plainer version of the handout's Figure 2, without the arrows. The
chunk has no `fig.cap`, so that the knitted sheet does not number it:
the figures numbered 1 to 3 in the PDF are your own, in Part 3.

```{r figure-2, fig.height = 3.2}
#ggplot(summers, aes(x = tmax_f, color = period)) +
#  geom_density(adjust = 1.5) +
#  labs(x = "Daily high (degrees F)", y = "Density", color = "Period")
```

The numbers the interpretation cites: the median of each period, and
the 10th and 90th percentiles of each.

```{r numbers-figure-2}
#median(early$tmax_f)
#median(late$tmax_f)
#quantile(early$tmax_f, c(0.1, 0.9))
#quantile(late$tmax_f, c(0.1, 0.9))
```

Please write the interpretation of Figure 2 in your own words: the
comparison, the numbers in their units, what they mean for the
question, and what the figure cannot establish. Every number is inline R.
The chunk below is not run; it only shows the form. A single number
from `quantile()` is written `quantile(late$tmax_f, 0.9)`, and the
shift is the difference of two such numbers.

```{r inline-form, eval = FALSE}
# the median day was `r median(late$tmax_f)` degrees F,
# against `r median(early$tmax_f)` degrees F in 1991-2005
```

*Your interpretation of Figure 2 here.*

## Figure 3: the hottest days

Figure 3 in the handout counts the days per summer at or above each
temperature. At 95 degrees F, the count for each period is the number
of such days divided by its fifteen summers.

```{r numbers-figure-3}
#sum(early$tmax_f >= 95) / 15
#sum(late$tmax_f >= 95) / 15
# How many times as common did those days become? Your line here.
```

*Two or three sentences that direct the reader to this feature of
Figure 3 and say what it means for the question, with the numbers as
inline R.*

## Three weak sentences

Each sentence below claims more than its figure shows. Please say
which words overstate the evidence, and rewrite the sentence to match
what the figure shows, with a number from the data where one is
needed.

> **1.** "Figure 1 proves that climate change is making Nashville's
> summers dramatically hotter."

*Your rewrite here.*

> **2.** "Figure 2 shows a significant difference between the two
> periods."

*Your rewrite here.*

> **3.** "Figure 3 shows that extreme heat is increasing over time."

*Your rewrite here.*

# Part 2: A Discussion for Nashville's summers

<!-- In your own words and with your own examples, not the handout's.
     Between 150 and 250 words in all is plenty. -->

Research question: *Are Nashville's summers getting hotter, and is it
the whole summer or only its hottest days?*

**Summary.** *Two or three sentences that answer both parts of the
question and need all three figures.*

**Implications.** *One implication: who would do what differently,
matched to what the figures show.*

**Limitations.** *Two limitations specific to this dataset, different
from the two in the handout.*

**Future directions.** *One next analysis that follows from one of
your limitations, specific enough that someone else could carry it
out.*

# Part 3: Your report

<!-- Your own dataset and your figures from Lab 3, or their current
     versions. Paste each figure's chunk into its section below, with
     its fig.cap. Under each figure, a chunk computes the numbers the
     paragraph cites, and the paragraph uses them as inline R. Please
     finish the first figure's paragraph before starting the second. -->

## The data

Please read your data file here, with the changes your Lab 2 card
recorded in field 8 (a sentinel value set to `NA` with `na_if()`, a
type converted).

```{r read-data}
#d <- read_csv("data/YOUR-FILE.csv")
```

**Research question, as it stands now:** *Your question here.*

## The first figure

```{r figure-one, fig.cap = "YOUR CAPTION: what is plotted, what one mark is, and the main pattern."}
# Paste the chunk for your first figure here.
```

```{r numbers-one}
# The numbers the paragraph cites. For a numeric variable compared
# across groups, for instance:
#d |>
#  group_by(GROUP_COLUMN) |>
#  summarize(n = n(),
#            median = median(COLUMN_ONE, na.rm = TRUE))
```

*Introduce: Figure 1 shows ... and it addresses ... of the question.*

*Direct: the feature to look at is ..., and the number that shows it
is ..., compared to ...*

*Interpret: which means, for the question, ...*

*What it cannot establish: ...*

## The second figure

```{r figure-two, fig.cap = "YOUR CAPTION."}
# Paste the chunk for your second figure here.
```

```{r numbers-two}
# The numbers the second paragraph cites.
```

*Introduce: Figure 2 builds on Figure 1 by ...*

*Direct: ...*

*Interpret: ...*

*What it cannot establish: ...*

## The third figure

```{r figure-three, fig.cap = "YOUR CAPTION."}
# Paste the chunk for your third figure here.
```

```{r numbers-three}
# The numbers the third paragraph cites.
```

*Introduce: Figure 3 ...*

*Direct: ...*

*Interpret: ...*

*What it cannot establish: ...*

## Discussion

<!-- Write the summary first. If it can be written from one of your
     figures alone, consider refining the question or replacing a
     figure so that all three contribute distinct findings. -->

**Summary.** *The answer to your research question, in two or three
sentences, needing all three figures.*

**Implications.** *Who would do what differently, matched to your
evidence.*

**Limitations.** *What is specific to your data: the sample, the
measurement, the design, the scope.*

**Future directions.** *The next analysis, following from a
limitation.*

# Generative AI disclosure

*Which tools you used on this lab and what for, or "none".*

*Your answer here.*
