---
title: "Lab 1: Introduction to Analysis Reports"
author: "FIRSTNAME LASTNAME"
date: "`r format(Sys.time(), 'Compiled %X on %A, %B %d, %Y')`"
output: pdf_document
---

<!--

This is a comment in RMarkdown. Nothing between the arrow markers
above and below this text reaches the knitted document, so this is
where the instructions for each step live. You can delete a comment
once you have done what it says, or leave it; either is fine.

Before you start: rename this file Lab-1_FIRSTNAME_LASTNAME.Rmd, put
your name in the YAML header above (keep the quotation marks), and
make sure data/penguins.csv is in a data/ folder beside this file.

The R Markdown cheatsheet, which Part 1 refers to:
https://rstudio.github.io/cheatsheets/html/rmarkdown.html

-->

# Part 1: RMarkdown

## Knitting, and knitting errors

RMarkdown is a notebook-like tool for writing text and code in one
document. The output can take many forms (the two most common are
HTML and PDF) and in this class we knit to PDF. A PDF is produced by
LaTeX, a typesetting program that, like R, has packages that extend
what it can do. Two of those come pre-loaded and let us write links.

The RMarkdown way to write a link is square brackets for the text the
reader sees, then parentheses for the address:

[TEXT HERE](URL HERE)

> **1.** Please make the line above a working link to Vanderbilt's
> Data Science Institute (https://www.vanderbilt.edu/datascience/)
> with the text reading "Vanderbilt Data Science".

The LaTeX way does the same thing with the `hyperref` package, and
lets you colour the text with `xcolor`:

\href{URL HERE}{\color{blue} TEXT HERE}

> **2.** Please do the same as in 1, but with the LaTeX form.

Now a case where the knit fails until you add a package. The line in
the comment below uses `\st{}`, a command from the `soul` package,
which is not loaded.

<!-- Move the next line out of this comment (put it on its own line
     below the arrow), then knit to PDF.

\st{This text should be struck through}

-->

> **3.** What did knitting stop with? Please paste the error message
> into the comment below, then move the `\st{}` line back inside its
> comment so the document knits again.

<!-- YOUR ERROR MESSAGE HERE -->

The message names the thing LaTeX did not recognise. One fix is to
add the package under `header-includes:` in the YAML header; the
other is to not use the command. Either way, the message told you
where to look.

Copying and pasting into RMarkdown is another way a knit fails. Many
of these failures are characters that look ordinary and are not. The
equation in the comment below was pasted from a chat with Claude.

<!-- Move the next line out of this comment, then knit to PDF.

MAE = (1/n) × Σ|yi - ŷi|

-->

> **4.** What is the error this time, specifically? Please paste it
> into the comment below, then put the equation back in its comment.

<!-- YOUR ERROR MESSAGE HERE -->

> **5.** Drop the same error message into a generative AI of your
> choice and ask what it means. In two or three sentences below, say
> what it told you and whether that matched what you had already
> worked out from the message itself.

*Your answer here.*

As a general guideline, knit often. It is better to run into a
problem early than to find six at the end. Debugging is part of
programming, and RMarkdown is a program you are writing.

<!--

                 ^   ^   ^   ^   ^
                /|\ /|\ /|\ /|\ /|\
               / | X | X | X | X | \
              |__|___|___|___|___|__|
              |                     |
              |_____________________|

                    KEEP CALM

                      AND

             READ THE ERROR MESSAGE

-->

## Code chunks

Code and its output sit in line with the text through code chunks
like the one below.

```{r}
cat("This is an R code chunk. You can tell by the ```r")
```

Each chunk has two buttons at its top right. The green arrow runs
this chunk. The grey arrow-to-bar runs every chunk above it first,
which you need when this chunk depends on an object an earlier chunk
made. Try the green arrow on the chunk above now.

### Chunk options

The options after the `r` inside the braces control what a chunk
does and what the reader sees. The chunk below spells out every
option at its default value.

```{r, eval = TRUE, echo = TRUE, warning = TRUE, error = FALSE, message = TRUE, tidy = FALSE, results = "markup", cache = FALSE, comment = "##"}
cat("Default R code chunk options")
```

The cheatsheet says what each one means. Please change the options
on the next four chunks so that each does what its instruction says.

> **6.** Make this chunk *run* but *not show* its code.

```{r, eval = TRUE, echo = TRUE, warning = TRUE, error = FALSE, message = TRUE, tidy = FALSE, results = "markup", cache = FALSE, comment = "##"}
cat("You can't see me -- John Cena")
```

> **7.** Make this chunk *show* its code but *not run*.

```{r, eval = TRUE, echo = TRUE, warning = TRUE, error = FALSE, message = TRUE, tidy = FALSE, results = "markup", cache = FALSE, comment = "##"}
cat("I'm here... but I'm not there")
```

> **8.** Make the output appear without the `##` in front of it (the
> value to use is `NA`).

```{r, eval = TRUE, echo = TRUE, warning = TRUE, error = FALSE, message = TRUE, tidy = FALSE, results = "markup", cache = FALSE, comment = "##"}
cat("When \"no comment\" shows up, it's usually because the truth is too busy laughing\n -- GPT-4o\n")
```

> **9.** Make this figure 5 inches *wide*, 3 inches *tall*,
> *centred* on the page, with the caption "Histogram of 100 random
> numbers from 1 to 10".

```{r}
hist(sample(x = 1:10, size = 100, replace = TRUE),
     main = "Histogram!", xlab = "RaNdOm NuMbErS")
```

The two chunks below are set up for you. The first shows its warning
and its message; the second has `warning = FALSE` and
`message = FALSE`.

```{r, warning = TRUE, message = TRUE}
cat("This chunk is so fresh and so clean")
warning("Warning! A warning is not an error; the code still ran.")
message("Message in a bottle.")
```

```{r, warning = FALSE, message = FALSE}
cat("This chunk is so fresh and so clean")
warning("Warning! You won't be able to see this warning, though.")
message("Message in a bottle... that won't be received")
```

> **10.** Knit, and compare the two. What happened to the output of
> the second chunk, and when would you want that in a report?

*Your answer here.*

### Code that runs off the page

A long line of code is printed as one long line, and a PDF page has
an edge. The chunk below is one line.

```{r}
average_body_mass_by_species <- c(Adelie = 3700.66, Chinstrap = 3733.09, Gentoo = 5076.02); average_body_mass_by_species
```

> **11.** Please rewrite the chunk so that no line is longer than
> about 70 characters and the result is the same. (RStudio draws a
> faint vertical line at 80 characters in the editor; staying left of
> it is the habit.) One way is to break the vector across lines after
> a comma; another is to put the assignment and the printing on
> separate lines.

## Equations

Dollar signs produce mathematics. One pair keeps it in the line, like
$x$, and two pairs put it on its own line:

$$x$$

Formal hypotheses are written this way:

$$
H_0: \mu = 0
$$

$$
H_A: \mu \neq 0
$$

A LaTeX chunk does the same with the `aligned` environment:

```{=tex}
\begin{equation}
\begin{aligned}
H_0: \mu = 0 \\
H_A: \mu \neq 0
\end{aligned}
\end{equation}
```

For the notation, see
https://www.overleaf.com/learn/latex/Mathematical_expressions
and, for the symbols,
https://oeis.org/wiki/List_of_LaTeX_mathematical_symbols.

> **12.** The mean of $n$ values $x_1, x_2, \ldots, x_n$ is their sum
> divided by $n$. Please write it as an equation, with a bar over the
> $x$ on the left, a fraction, and a summation sign with its limits
> ($i = 1$ underneath, $n$ on top).

$$
YOUR \ EQUATION \ HERE
$$

# Part 2: Introduction to analysis reports

<!-- These questions are about the handout. A sentence or two each,
     in your own words, is what is wanted. -->

> **1.** In your own words, what is the purpose of the analysis
> reports?

*Your answer here.*

> **2.** What are the four skills the reports are building, and which
> two are the emphasis?

*Your answer here.*

> **3.** There is a subjective impression given on every report that
> does not count for points. Who gives it, from what perspective, and
> why bother?

*Your answer here.*

> **4.** The syllabus gives a two-part philosophy on generative AI. Do
> you agree or disagree with it, and why?

*Your answer here.*

> **5.** Which functions print output that you should *not* leave in
> an analysis report?

*Your answer here.*

> **6.** You have 10 plots with the same $x$-axis. What functions turn
> them into a single figure?

*Your answer here.*

> **7.** What is the class's stance on pre-grading?

*Your answer here.*

# Part 3: A dataset card for `penguins.csv`

<!-- The nine fields are those in Part 3 of the handout. The R chunks
     for fields 5, 6, and 8 are handed to you commented out; remove
     the hash characters when the CSV is in place. Fields 1 to 4, 7,
     and 9 come from the documentation:

       ?penguins in the R Console (R 4.5 or newer)
       https://allisonhorst.github.io/palmerpenguins/
       the three Environmental Data Initiative records whose DOIs are
       on the help page

     Where a field cannot be answered from the documentation, say so
     in the field. "The source does not state a licence" is an
     answer; a blank is not. -->

First read the file and look at what R made of it.

```{r read-penguins}
#penguins <- read.csv("data/penguins.csv")
#str(penguins)
```

## 1. Name and source

*What the dataset is called, where you obtained it (a readable link,
not a bare URL), who posted it there, and the date you downloaded it.
For this lab the CSV came from Brightspace, and I wrote it out of the
copy that ships with R. Say both.*

*Your answer here.*

## 2. Collector and purpose

*Who originally collected the data, and what for. The help page's
References section and the palmerpenguins website both name the
study.*

*Your answer here.*

## 3. Collection

*How, when, and where the penguins were measured, and what one row of
the file represents.*

*Your answer here.*

## 4. Licence and permissions

*What the source says you may do with the data. Look on the
palmerpenguins website and on the Environmental Data Initiative
record.*

*Your answer here.*

## 5. Structure

The size of the table and the names of its columns:

```{r size}
#nrow(penguins)
#ncol(penguins)
#names(penguins)
```

`unique()` gives the distinct values in a column, and `table()`
counts how many rows take each value. `range()` gives the smallest
and largest value of a numeric column, and needs `na.rm = TRUE` for
the same reason `mean()` did on Thursday.

```{r levels-and-ranges}
#unique(penguins$species)
#table(penguins$island)
#table(penguins$sex)
#range(penguins$bill_length_mm, na.rm = TRUE)
#range(penguins$body_mass_g, na.rm = TRUE)
#range(penguins$year)
```

Please complete the table. The first two rows are done. "Stored as"
is what `str()` reported, not what the documentation says the column
is; where the two differ, the last column of the table is the place
to say so.

| Column              | Meaning                              | Stored as | Units | Range or levels                    |
|:----------------------------|:--------------------------------|:---------|:------|:------------------------------------|
| `species`           | Penguin species                      | chr       |       | Adelie, Chinstrap, Gentoo          |
| `island`            | Island in the Palmer Archipelago     | chr       |       | Biscoe, Dream, Torgersen           |
| `bill_length_mm`    |                                      |           |       |                                    |
| `bill_depth_mm`     |                                      |           |       |                                    |
| `flipper_length_mm` |                                      |           |       |                                    |
| `body_mass_g`       |                                      |           |       |                                    |
| `sex`               |                                      |           |       |                                    |
| `year`              |                                      |           |       |                                    |

> The documentation says `species`, `island`, and `sex` are factors.
> `str()` says something else. Why do they differ, and which one is
> right about the file you have?

*Your answer here.*

## 6. Missing values

`is.na()` returns `TRUE` wherever a value is missing. Adding up
`TRUE`s counts them, and `colSums()` does that one column at a time.

```{r missing}
#colSums(is.na(penguins))
#penguins[is.na(penguins$sex), ]
```

*How is a missing value written in the CSV file itself (open it in a
text editor and look)? How many are there, in which columns, and is
there a pattern to which rows they fall in?*

*Your answer here.*

## 7. Known issues

*Caveats from the documentation, and anything you noticed yourself
that a reader should know before trusting a figure drawn from this
file.*

*Your answer here.*

## 8. Changes you made

*Anything you did to the data between the source and here. If you
made no changes, say so. If you did make one, the code for it belongs
here too.*

```{r changes}
# Any changes to the data go in this chunk, so the reader can see them.
```

*Your answer here.*

## 9. Citation

*How the source asks to be cited, and the citation itself, formatted
as in the handout's reading boxes.*

*Your answer here.*

# Generative AI disclosure

*Which tools you used on this lab and what for, or "none".*

*Your answer here.*
