---
title: "Unit 2 | Data Visualization"
author: "FIRSTNAME LASTNAME"
date: "`r format(Sys.time(), 'Compiled %X on %A, %B %d, %Y')`"
output: pdf_document
---

# Create a good plot and the worst plot ever

Come up with a survey question to ask the class and add it to our Microsoft Forms: https://forms.office.com/Pages/DesignPageV2.aspx?subpage=design&FormId=OX9aur7js0q0UGf6gPrsrT_5kjO3ittLkBAIvlszyF5UNkpLR1pORThGTEJDQTVZMzBBQVJEQkxYQy4u&Token=04b2d16ec09f4b1e9ca2a7f4488d538c

# \{ggplot2\}

## Load \{ggplot2\} and `PluckData.RData`

```{r load ggplot2, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Load packages
library(ggplot2); library(tidyverse); library(GGally)

# Load .RData
load("./Unit_02/data/PluckData.RData")

```

# Univariate

## Histogram

```{r str example, echo = FALSE, eval = FALSE, comment = NA, warning = FALSE, message = FALSE}
# Base R
str(dataset)
```

```{r base R histogram, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Base R
hist(dataset$GPA, breaks = 10)
```

Something is very wrong here... Figure out what's wrong and re-plot it

```{r base R histogram fixed, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
summary(dataset$GPA)
range(dataset$GPA, na.rm = TRUE)
```

<!--- update to `dataset_clean` --->

```{r base R histogram fixed, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
dataset_clean <- dataset

dataset_clean$GPA[
  dataset_clean$GPA > 5
] <- NA

range(dataset_clean$GPA, na.rm = TRUE)
```

## Customize

```{r base R color, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Customized
hist(
  dataset_clean$GPA, 
  col = "hotpink", # color
  xlab = "GPA", # x-axis label
  main = "Distribution of GPA", # title
  breaks = 10, # number of bars
  xlim = c(0, 4) # restricting range to all possible values
)
```

## Boxplot

```{r base R boxplot, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Customized
boxplot(
  dataset_clean$GPA, 
  col = "gold", # color
  xlab = "GPA", # x-axis label
  main = "Boxplot of GPA", # title
  ylim = c(0, 4) # restricted range of possible values
)
```

Let's do multiple

```{r base R boxplot multiple, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Customized
boxplot(
  dataset_clean[,7:8], 
  col = c("orange", "gold"), # color
  xlab = "Personality", # x-axis label
  main = "Boxplot of Personality Traits", # title
  horizontal = TRUE # flip to make along y-axis
)
```

## Descriptives

```{r base R barplot descriptives, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Customized
barplot(
  table(dataset_clean$Sex), 
  col = c("lightblue", "lightpink"),
  xlab = "Sex"
)
```

# \{ggplot2\}

Base R plots aren't the most aesthetically pleasing but they're quick and good for checking your data before jumping into more time-intensive plotting like \{ggplot2\}

```{r ggplot2 histogram, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}
ggplot( # base function
  data = dataset_clean, # data to be used *across* all layers
  aes( # sets the "looks" of the plot
    x = GPA # put GPA on the x-axis
  )
) + # "add" a layer
  geom_histogram( # add histogram graphic
    bins = 10 # set number of "bins" to 10
  )
```

Let's "re-create" base R's plot with \{ggplot2\}

```{r ggplot2-to-base histogram, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}
ggplot( # base function
  data = dataset_clean, # data to be used *across* all layers
  aes( # sets the "looks" of the plot
    x = GPA # put GPA on the x-axis
  )
) + # "add" a layer
  geom_histogram( # add histogram graphic
    color = "black", # color of the *outline* of the bars
    fill = "lightgrey", # color of the inside of the bars
    bins = 10 # set number of "bins" to 10
  ) + # "add" a layer
  labs( # change labels
    y = "Frequency", title = "Histogram"
  ) + # "add" a layer
  theme( # the "style" of your plot
    panel.background = element_blank(), # takes away ALL of the background
    axis.line = element_line(linewidth = 0.50, lineend = "square"), # adds axis line
    axis.text = element_text(size = 10), # changes size
    axis.title = element_text(size = 12), # changes size (should usually be larger than text)
    plot.title = element_text(
      size = 14, hjust = 0.50 # adjusts horizontal position
    )
  )
```

Whoa... that's a lot just to match base R. The benefit of \{ggplot2\} is the flexibility and professional-looking plots. You can also store parts of plots as objects. For example, I use the above `theme` a lot. Let's store it as `my_theme`

```{r my_theme2, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Store theme
my_theme <- theme( # the "style" of your plot
  panel.background = element_blank(), # takes away ALL of the background
  axis.line = element_line(linewidth = 0.50, lineend = "square"), # adds axis line
  axis.text = element_text(size = 10), # changes size
  axis.title = element_text(size = 12), # changes size (should usually be larger than text)
  plot.title = element_text(
    size = 14, hjust = 0.50 # adjusts horizontal position
  )
)
```

## Aesthetics

### Fill

```{r my_theme, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}
ggplot( # base function
  data = dataset_clean, # data to be used *across* all layers
  aes( # sets the "looks" of the plot
    x = GPA # put GPA on the x-axis
  )
) + # "add" a layer
  geom_histogram( # add histogram graphic
    color = "black", # color of the *outline* of the bars
    fill = "hotpink", # color of the inside of the bars
    bins = 10 # set number of "bins" to 10
  ) + # "add" a layer
  labs( # change labels
    y = "Frequency", title = "Histogram"
  ) + my_theme
```

### Color

```{r my_theme4, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}
ggplot( # base function
  data = dataset_clean, # data to be used *across* all layers
  aes( # sets the "looks" of the plot
    x = GPA # put GPA on the x-axis
  )
) + # "add" a layer
  geom_histogram( # add histogram graphic
    color = "lightgray", # color of the *outline* of the bars
    fill = "#FFBF69", # color of the inside of the bars
    bins = 10 # set number of "bins" to 10
  ) + # "add" a layer
  labs( # change labels
    y = "Frequency", title = "Histogram"
  ) + my_theme
```

### Color

```{r my_theme3, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}
ggplot( # base function
  data = dataset_clean, # data to be used *across* all layers
  aes( # sets the "looks" of the plot
    x = GPA # put GPA on the x-axis
  )
) + # "add" a layer
  geom_histogram( # add histogram graphic
    color = "black", # color of the *outline* of the bars
    fill = "lightgray", # color of the inside of the bars
    bins = 4 # set number of "bins" to 10
  ) + # "add" a layer
  labs( # change labels
    y = "Frequency", title = "Histogram"
  ) + my_theme
```

## Density

Now, let's do a density plot with and without `my_theme`. First, let's store the plot *without* the theme as `my_density`

```{r density without theme, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}
my_density <- ggplot( # base function
  data = dataset_clean, # data to be used *across* all layers
  aes( # sets the "looks" of the plot
    x = GPA # put GPA on the x-axis
  )
) + # "add" a layer
  geom_density(
    color = "black", # color of the *outline*
    fill = "lightgrey", # color of the underneath area
  ) +
  geom_boxplot(width = 0.05,  fill = "lightgrey", aes(y = -0.05)) +
  my_theme +
  labs(y = "Density")

# Plot it
my_density
```

Add the `my_theme` layer to `my_density`

```{r density with theme, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}

```

Let's also change the `limits` of the plot to plot the full range of possible GPAs

```{r density limits, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}
my_density +
  scale_x_continuous(limits = c(0.00, 4.00))
```

We can also add `breaks` to align with different letter grades

```{r density breaks, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# {ggplot2}
my_density +
  scale_x_continuous(
    name = "Letter Grade",
    limits = c(0.00, 4.00), breaks = round(
      seq(0.333, 4.00, 1/3), 3
      # sequence and round to get appropriate breaks
    ),
    labels = c("F", "D-", "D", "D+", "C-", "C", "C+", "B-", "B", "B+", "A-", "A")
  ) +
  # theme_minimal() +
  theme(
    axis.text = element_text(size = 16),
    # axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 18)
  )
```

## Layering

```{r layered one, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Create x-axis object
x_axis <- scale_x_continuous(
    limits = c(0.00, 4.00), breaks = round(seq(0.00, 4.00, 1/3), 3)
  )

# Create histogram as *first* layer
ggplot(data = dataset, aes(x = GPA)) +
  geom_histogram( # first layer
    color = "black", fill = "lightgrey", bins = 10 
  ) +
  geom_boxplot( # second layer
    color = "black", fill = "lightgrey"
  ) + x_axis + my_theme

# Hmm... we need to move the boxplot
```


```{r lower boxplot, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Create histogram as *first* layer
ggplot(data = dataset, aes(x = GPA)) +
  geom_histogram( # first layer
    color = "black", fill = "lightgrey", bins = 10 
  ) +
  geom_boxplot( # second layer
    color = "black", fill = "lightgrey",
    aes(y = -6)
  ) + x_axis + my_theme
```

Use `?geom_boxplot` and find how to make the boxplot *wider* (set the value to 6)

```{r wider boxplot, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
ggplot(data = dataset, aes(x = GPA)) +
  geom_histogram( # first layer
    color = "black", fill = "lightgrey", bins = 10 
  ) +
  geom_boxplot( # second layer
    color = "black", fill = "lightgrey",
    aes(y = -6),
    width = 6
  ) + x_axis + my_theme
```

Now, we'll add points 

```{r points, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
x_axis <- scale_x_continuous(
    name = "Letter Grade",
    limits = c(0.00, 4.00), breaks = round(
      seq(0.333, 4.00, 1/3), 3
      # sequence and round to get appropriate breaks
    ),
    labels = c("F", "D-", "D", "D+", "C-", "C", "C+", "B-", "B", "B+", "A-", "A")
  )

my_theme <- theme(
    axis.text = element_text(size = 16),
    # axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 18)
  )


# Create histogram as *first* layer
ggplot(data = dataset, aes(x = GPA)) +
  geom_histogram( # first layer
    color = "black", fill = "lightgrey", bins = 10 
  ) +
  geom_boxplot( # second layer
    color = "black", fill = "lightgrey",
    aes(y = -4), width = 3
  ) + 
  geom_point( # third layer
    aes(y = -4), color = "black" # notice we used `color` rather than `fill`
  ) + x_axis + my_theme +
  labs(y = "Frequency", title = "Letter Grade Distribution") +
  theme(
    plot.title = element_text(
      size = 20, hjust = 0.50
    )
  )
```

Notice how the dots are in front of the boxplot? Let's put them behind the boxplot (hint: think about the order of the `geom_`)

```{r points behind, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
ggplot(data = dataset, aes(x = GPA)) +
  geom_histogram( # first layer
    color = "black", fill = "lightgrey", bins = 10 
  ) +
  geom_point( # third layer
    aes(y = -4), color = "black" # notice we used `color` rather than `fill`
  ) +
  geom_boxplot( # second layer
    color = "black", fill = "lightgrey",
    aes(y = -4), width = 3
  ) + x_axis + my_theme +
  labs(y = "Frequency", title = "Letter Grade Distribution") +
  theme(
    plot.title = element_text(
      size = 20, hjust = 0.50
    )
  )
```

To get fancy, ask generative AI to help you do the following:

+   Set the fill of `geom_boxplot` to be transparent

+   Remove the outliers from `geom_boxplot` so that they do not appear on the plot

+   Increase the spread of `geom_point` so that it acts as a "fill" of `geom_boxplot`

+   Change `geom_histogram` to `geom_density`

+   Change the `fill` of `geom_density` to `"#8CC7A1"`

+   Also, change the `color` of `geom_point` to the same color

+   Change the y-axis label to "Density"

+   Change `y` for both `geom_boxplot` and `geom_point` to `-0.10` and adjust settings for these `geom_*` as necessary

```{r raincloud plot, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
ggplot(data = dataset, aes(x = GPA)) +
  geom_density( # first layer
    color = "black", fill = "#8CC7A1", bins = 10 
  ) +
  geom_point( # third layer
    aes(y = -0.05), color = "#8CC7A1", # notice we used `color` rather than `fill`
    position = position_jitter(height = 0.05)
  ) +
  geom_boxplot( # second layer
    color = "black", fill = NA,
    aes(y = -0.05), width = 0.10, outlier.shape = NA
  ) + x_axis + my_theme +
  labs(y = "Frequency", title = "Letter Grade Distribution") +
  theme(
    plot.title = element_text(
      size = 20, hjust = 0.50
    )
  )
```

## Multiple Plots in One

```{r multiple plots, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Plot multiple based on grouping variable
ggplot(data = dataset_clean, aes(x = GPA)) +
  facet_wrap(~Sex) + # plot *by* group
  geom_boxplot() + x_axis + my_theme
```

```{r multiple plots fixed, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Subsetting missing data out
ggplot(data = dataset_clean, aes(x = GPA)) +
  facet_wrap(~Sex) +
  geom_boxplot(fill = c("lightblue", "lightpink")) + x_axis + my_theme + 
  coord_flip() + # flip x- and y-axis
  theme( # remove x-axis text and ticks
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

```

# Bivariate

Rather than splitting by a variable, we'll instead plot two (or more) variables together on the same plot

```{r two density, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Set data
ggplot(data = dataset_clean) + # notice there is no `aes`
  geom_density(aes(x = Extraversion), fill = "#FF99C9", color = "#FF99C9") +
  geom_density(aes(x = Neurotocism), fill = "#A2C7E5", color = "#A2C7E5") +
  my_theme
```


Hmm... the fill and colors did not abide by my selection. Let's try to fix that...

```{r two density colors, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Set data
ggplot(data = dataset) + # notice there is no `aes`
  geom_density(aes(x = Extraversion) , fill = "#478978", color = "#478978") +
  geom_density(aes(x = Neurotocism) , fill = "#A2C7E5", color = "#A2C7E5") +
  my_theme
```

Colors are working but we can't see the full `Extraversion` density

```{r two density colors2, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Set data
ggplot(data = dataset) + # notice there is no `aes`
  geom_density(aes(x = Extraversion) , fill = "#478978", color = "#478978", alpha = 0.50) +
  geom_density(aes(x = Neurotocism) , fill = "#A2C7E5", color = "#A2C7E5", alpha = 0.50) +
  my_theme
```

Looking good, for the most part, but it would look a lot better if we had a legend. For that, we'll need to do some data wrangling:

+   Use `pivot_longer`

    +   Get `cols = c("Extraversion", "Neurotocism")`
    
    +   Set `names_to = "Personality"`
    
    +   Set `values_to = "Values"`
    
+   Store as `dataset_long`
    
```{r two density wrangling, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}

```

Now, use these new variables to plot:

+   You'll only need *one* `geom_density` (set its `alpha = 0.50`)

+   You can set: `aes(x = Values, color = Personality, fill = Personality)`

+   Add the following: `scale_fill_manual(values = c("#478978", "#A2C7E5"))`

```{r two density plot, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}

```

Let's also take a look at boxplots:

+   You'll only need *one* `geom_boxplot`

+   Keep: `scale_fill_manual(values = c("#478978", "#A2C7E5"))`

```{r two boxplots, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}

```

## Scatterplot

Boxplots and density plots allow for the overlap of two distributions to be compared but don't tell us much about the *relationship* between two variables. For that, we can use a scatterplot

```{r base R scatter, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
plot(
  dataset$Extraversion, dataset$Neurotocism, pch = 19, # changes the type of point
  xlab = "Extraversion", ylab = "Neurotocism", main = "Personality",
  xlim = c(0, 25), ylim = c(0, 25) # changes the range of the axes
)
```

Re-creating in \{ggplot2\}

```{r ggplot2 scatter, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
ggplot(
  data = dataset, aes(x = Extraversion, y = Neurotocism)
) +
  geom_point() + my_theme +
  scale_x_continuous(limits = c(0, 25)) +
  scale_y_continuous(limits = c(0, 25))
```

Let's add a line to visualize the slope (direction) of the relationship

```{r ggplot2 line, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
ggplot(
  data = dataset, aes(x = Extraversion, y = Neurotocism)
) +
  geom_point() + my_theme +
  scale_x_continuous(limits = c(0, 25)) +
  scale_y_continuous(limits = c(0, 25)) +
  geom_smooth(method = "lm", se = FALSE)
```

Check to see whether there is a different relationship (slope) for `Sex` (hint: plot using a single plot)

```{r ggplot2 line2, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
ggplot(
  data = dataset[!is.na(dataset$Sex),], aes(x = Extraversion, y = Neurotocism)
) +
  geom_point() + my_theme +
  scale_x_continuous(limits = c(0, 25)) +
  scale_y_continuous(limits = c(0, 25)) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Sex)
```

## `ggpairs`

The \{GGally\} package has an amazing function called `ggpairs` that will streamline all of these visuals into a single plot (only use numeric variables!)

```{r ggpairs, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
ggpairs(dataset[,c(3, 5:8)])
```

# Create a good plot and the worst plot ever

1.    With your table, come up with a research question, based on the survey questions from across the class, to visualize

2.    Create your own good plot and the worst plot ever to answer the research question

3.    Submit to Brightspace